---
title: "Jane Austen Hapax"
author: "Stephen Kaluzny"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: yes
    theme: cerulean
---
# Introduction
A _hapax_ is (or _hapax legomenon_) is a word that only occurs once in
a work or collection of works.
See https://en.wikipedia.org/wiki/Hapax_legomenon

How common are hapaxes in works of literature?
With access to digital versions of the works,
finding hapaxes should be a straighforward computational task.

This paper considers hapaxes in the works by the Jane Austen.
Austen's works are all available from Project Gutenberg
(https://en.wikipedia.org/wiki/Project_Gutenberg)

The R CRAN package `gutenburgr` provides an API to works from
Project Gutenberg.
The R CRAN package `tidytext` provides functions to process the text.

## Setup

The packages we use.
```{r}
library(dplyr, quietly=TRUE, warn.conflicts=FALSE)
library(ggplot2, quietly=TRUE, warn.conflicts=FALSE)
library(gutenbergr)
library(tidytext)
```

# Data
Get the works of Jane Austen from Project Gutenberg.

```{r}
austen_works <- gutenberg_works(author == "Austen, Jane")
```

```{r}
nrow(austen_works)
names(austen_works)
austen_works[["title"]]
```

Only look at individual Austen's novels:

```{r only_novels}
austen_works <- austen_works[-c(9:10),]
austen_works[["title"]]
```

```{r download, eval=FALSE}
austen_books <- gutenberg_download(austen_works[["gutenberg_id"]],
  meta_fields="title",
  mirror="http://mirrors.xmission.com/gutenberg/")
```

```{r load_or_download, echo=FALSE}
if(file.exists(here::here("data", "austen_books.rds"))) {
  austen_books <- readRDS(here::here("data", "austen_books.rds"))
} else {
  austen_books <- gutenberg_download(austen_works[["gutenberg_id"]],
    meta_fields="title",
    mirror="http://mirrors.xmission.com/gutenberg/")
  saveRDS(austen_books, here::here("data", "austen_books.rds"))
}
```

```{r}
nrow(austen_books)
names(austen_books)
austen_books
```

# Word Analysis

Get words in each book,
drop any formatting characters (e.g. \_) around the words.
This produces a `tibble` with each observation (row)
containg the book title and the word.
```{r words}
words <- austen_books |>
  unnest_tokens(word, text) |>
  mutate(word = stringr::str_extract(word, "[a-z']+"))
```

```{r}
nrow(words)
words
```

Count words per book, dropping "stop" words first.
```{r count_words}
book_word_counts <- words |>
  anti_join(stop_words, by="word") |>
  count(title, word, sort=TRUE)
```

```{r} 
book_word_counts
```
## Words Per Book
```{r}
words_per_book <- book_word_counts |>
  group_by(title) |>
  summarise(total_words = sum(n), unique_words = length(word))
```

```{r}
words_per_book
```

## Hapax in Each Book

Hapax (count == 1) words by book:
```{r rare_words}
book_hapax_words <- book_word_counts |>
  filter(n == 1) |>
  arrange(word)
```

```{r}
nrow(book_hapax_words)
book_hapax_words
```
  
Number of hapaxes per book,
combined with total number of words and number of unique words.

```{r combine_by_book}
book_hapax_words <- book_hapax_words |> group_by(title) |>
  summarise(n_hapax = n()) |>
  ungroup() |>
  left_join(words_per_book, by="title") |>
  mutate(percent_hapax_total = n_hapax / total_words * 100,
    percent_hapax_unique = n_hapax / unique_words * 100)
```
```{r}
knitr::kable(book_hapax_words, format.args=list(digits = 4))
```

```{r hapax_per_book}
book_hapax_words |> group_by(title) |>
  summarise(n_hapax = n()) |>
  arrange(desc(n_hapax)) |>
  ggplot(aes(y=factor(title, levels=title), x=n_hapax)) +
    geom_point() +
    geom_segment(aes(x=0, xend=n_hapax, y=title, yend=title)) +
    ggtitle("Number of Hapaxes in Each Austen Novel") +
    ylab("Title") +
    xlab("Number of Hapaxes")
```

## Hapax for All Austen Works

Hapax for all Austen works.
```{r hapax_all_works}
works_hapax_words <- words |>
  anti_join(stop_words, by="word") |>
  count(word, sort=TRUE) |>
  filter(n < 2)
```

```{r}
# Number of hapaxes for all of Austen works
nrow(works_hapax_words)
```

A sample of hapaxes across all of Austen's works:from Austen's works:
```{r sample_hapaxes}
set.seed(42)
sample(works_hapax_words[["word"]], 50)
```
