---
title: "Charles Dickens Hapax"
author: "Stephen Kaluzny"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: yes
    theme: cerulean
---
# Introduction
A _hapax_ is (or _hapax legomenon_) is a word that only occurs once in
a work or collection of works.
See https://en.wikipedia.org/wiki/Hapax_legomenon

How common are hapaxes in works of literature?
With access to digital versions of the works,
finding hapaxes should be a straighforward computational task.

This paper considers hapaxes in the works by the Charles Dickens.
Dickens's works are all available from Project Gutenberg
(https://en.wikipedia.org/wiki/Project_Gutenberg)

The R CRAN package `gutenburgr` provides an API to works from
Project Gutenberg.
The R CRAN package `tidytext` provides functions to process the text.

## Setup

The packages we use.
```{r}
library(dplyr, quietly=TRUE, warn.conflicts=FALSE)
library(ggplot2, quietly=TRUE, warn.conflicts=FALSE)
library(gutenbergr)
library(tidytext)
```

# Data
Get the works of Charles Dickens from Project Gutenberg.

```{r}
dickens_works <- gutenberg_works(author == "Dickens, Charles")
```

Novels by Dickens (from Wikipedia):
```{r novels}
dickens_novel_titles <- scan(here::here("data", "dickens_novels.txt"),
  what="", sep="\n")
```

Novels from Project Gutenberg.
```{r}
indx_novels <- match(dickens_novel_titles, dickens_works[["title"]])
missing_novels <- dickens_novel_titles[is.na(indx_novels)]
indx_novels <- indx_novels[!is.na(indx_novels)]
indx_missing <- sapply(missing_novels,
  function(x) grep(x, dickens_works[["title"]]))
indx_novels <- c(indx_novels, indx_missing)
dickens_works[indx_novels, "title"]
```

Download Dickens novels from Project Gutenberg.
```{r download, eval=FALSE}
dickens_books <- gutenberg_download(dickens_works[indx_novels, "gutenberg_id"],
  meta_fields="title",
  mirror="http://mirrors.xmission.com/gutenberg/")
```

```{r load_or_download, echo=FALSE}
if(file.exists(here::here("data", "dickens_books.rds"))) {
  dickens_books <- readRDS(here::here("data", "dickens_books.rds"))
} else {
dickens_books <- gutenberg_download(dickens_works[indx_novels, "gutenberg_id"],
  meta_fields="title",
  mirror="http://mirrors.xmission.com/gutenberg/")
  saveRDS(dickens_books, here::here("data", "dickens_books.rds"))
}
```

```{r}
nrow(dickens_books)
names(dickens_books)
dickens_books
```

We shorten some of the longer titles to make displaying of results better.

```{r title_shorten}
dickens_books[["title"]] <- ifelse(
  dickens_books[["title"]] == "The Haunted Man and the Ghost's Bargain",
  "The Haunted Man", dickens_books[["title"]])
dickens_books[["title"]] <- ifelse(
  dickens_books[["title"]] == "Barnaby Rudge: A Tale of the Riots of 'Eighty",
  "Barnaby Rudge", dickens_books[["title"]])
dickens_books[["title"]] <- ifelse(
  dickens_books[["title"]] == "The Chimes\r\nA Goblin Story of Some Bells That Rang an Old Year out and a New Year In",
  "The Chimes", dickens_books[["title"]])
```

# Word Analysis

Get words in each book,
drop any formatting characters (e.g. \_) around the words.
This produces a `tibble` with each observation (row)
containg the book title and the word.
```{r words}
words <- dickens_books |>
  unnest_tokens(word, text) |>
  mutate(word = stringr::str_extract(word, "[a-z']+"))
```

```{r}
nrow(words)
words
```

Count words per book, dropping "stop" words first.
```{r count_words}
book_word_counts <- words |>
  anti_join(stop_words, by="word") |>
  count(title, word, sort=TRUE)
```

```{r}
book_word_counts
```

## Words Per Book
```{r}
words_per_book <- book_word_counts |>
  group_by(title) |>
  summarise(total_words = sum(n), unique_words = length(word))
```

```{r}
words_per_book
```

## Hapax in Each Book

Hapax (count == 1) words by book:
```{r rare_words}
book_hapax_words <- book_word_counts |>
  filter(n == 1) |>
  arrange(word)
```

```{r}
nrow(book_hapax_words)
book_hapax_words
```
  
Number of hapax per book,
combined with total number of words and number of unique words.

```{r combine_by_book}
book_hapax_words <- book_hapax_words |> group_by(title) |>
  summarise(n_hapax = n()) |>
  ungroup() |>
  left_join(words_per_book, by="title") |>
  mutate(percent_hapax_total = n_hapax / total_words * 100,
    percent_hapax_unique = n_hapax / unique_words * 100)
```
```{r}
knitr::kable(book_hapax_words, format.args=list(digits = 4))
```

```{r hapax_per_book}
book_hapax_words |> group_by(title) |>
  arrange(desc(n_hapax)) |>
  ggplot(aes(y=factor(title, levels=title), x=n_hapax)) +
    geom_point() +
    geom_segment(aes(x=0, xend=n_hapax, y=title, yend=title)) +
    ggtitle("Number of Hapaxes in Each Dickens Novel") +
    ylab("Title") +
    xlab("Number of Hapaxes")
```

```{r hapax_percent_words}
book_hapax_words |> group_by(title) |>
  arrange(desc(n_hapax)) |>
  ggplot(aes(y=factor(title, levels=title), x=percent_hapax_total)) +
    geom_point() +
    geom_segment(aes(x=0, xend=percent_hapax_total, y=title, yend=title)) +
    ggtitle("Percent Hapaxes of Words in Each Dickens Novel") +
    ylab("Title") +
    xlab("Percent Hapaxes of Words")
```

```{r hapax_percent_unique_words}
book_hapax_words |> group_by(title) |>
  arrange(desc(n_hapax)) |>
  ggplot(aes(y=factor(title, levels=title), x=percent_hapax_unique)) +
    geom_point() +
    geom_segment(aes(x=0, xend=percent_hapax_unique, y=title, yend=title)) +
    ggtitle("Percent Hapaxes of Unique Words in Each Dickens Novel") +
    ylab("Title") +
    xlab("Percent Hapaxes of Unique Words")
```

## Hapax for All Dickens Works

Hapax for all Dickens works.
```{r hapax_all_works}
works_hapax_words <- words |>
  anti_join(stop_words, by="word") |>
  count(word, sort=TRUE) |>
  filter(n < 2)
```

```{r}
# Number of hapaxes for all of Dickens works
nrow(works_hapax_words)
```

A sample of hapaxes across all of Dickens's works:from Dickens's works:
```{r sample_hapaxes}
set.seed(42)
sample(works_hapax_words[["word"]], 50)
```
